---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# INLAspacetime

<!-- badges: start -->
[![CRAN Status](http://www.r-pkg.org/badges/version-last-release/INLAspacetime)](https://cran.r-project.org/package=INLAspacetime)
[![](https://cranlogs.r-pkg.org/badges/INLAspacetime)](https://cran.r-project.org/package=INLAspacetime)
[![check no-suggestions](https://github.com/eliaskrainski/INLAspacetime/workflows/R-CMD-check-no-suggests/badge.svg)](https://github.com/eliaskrainski/INLAspacetime/actions)
[![check](https://github.com/eliaskrainski/INLAspacetime/workflows/R-CMD-check/badge.svg)](https://github.com/eliaskrainski/INLAspacetime/actions)
[![pkgdown](https://github.com/eliaskrainski/INLAspacetime/workflows/pkgdown/badge.svg)](https://github.com/eliaskrainski/INLAspacetime/actions)
<!-- badges: end -->

This is a R package to implement certain spatial and spatio-temporal models,
including some of the spatio-temporal models proposed 
[here](https://www.idescat.cat/sort/sort481/48.1.1.Lindgren-etal.pdf). 
It uses the `cgeneric` interface in the INLA package, 
to implement models by writing `C`
code to build the precision matrix compiling it so that 
INLA can use it internally.

## We have implemented 

1. some of the models presented in 
https://www.idescat.cat/sort/sort481/48.1.1.Lindgren-etal.pdf

2. the barrier model proposed in 
https://doi.org/10.1016/j.spasta.2019.01.002

## Vignettes

Please chek [Tutorials](https://eliaskrainski.github.io/INLAspacetime/)

## Installation

The 'INLA' package is a suggested one, but you will need it 
for actually fitting a model. You can install it with
```{r eval = !require(INLA)}
install.packages("INLA",repos=c(getOption("repos"),INLA="https://inla.r-inla-download.org/R/testing"), dep=TRUE) 
```


You can install the current [CRAN](https://CRAN.R-project.org) version of INLAspacetime:
```{r cran-installation, eval = FALSE}
install.packages("INLAspacetime")
```

You can install the latest version of INLAspacetime from [GitHub](https://github.com/eliaskrainski/INLAspacetime) with
```{r ghinst, eval = FALSE}
## install.packages("remotes")
remotes::install_github("eliaskrainski/INLAspacetime",  build_vignettes=TRUE)
```
<!-- or track the development version builds via [inlabru-org.r-universe.dev](https://inlabru-org.r-universe.dev/ui#builds): -->
<!-- ```{r universe-installation, eval = FALSE} -->
<!-- ## Enable universe(s) by inlabru-org -->
<!-- options(repos = c( -->
<!--   inlabruorg = "https://inlabru-org.r-universe.dev", -->
<!--   INLA = "https://inla.r-inla-download.org/R/testing", -->
<!--   CRAN = "https://cloud.r-project.org" -->
<!-- )) -->

<!-- ## Install it -->
<!-- install.packages("INLAspacetime") -->
<!-- ``` -->

# A spacetime example

Simulate some fake data.

```{r fakedata}
set.seed(1)
n <- 5
dataf <- data.frame(
    s1   = runif(n, -1, 1),
    s2   = runif(n, -1, 1),
    time = runif(n, 1, 4),
    y    = rnorm(n, 0, 1))
str(dataf)
```

Loading packages:
```{r library}
library(fmesher)
library(INLA)
library(INLAspacetime)
```

Define spatial and temporal discretization meshes
```{r meshes}
smesh <- fm_mesh_2d(
  loc = cbind(0,0), 
  max.edge = 5, 
  offset = 2)
tmesh <- fm_mesh_1d(
  loc = 0:5)
```

Define the spacetime model object to be used
```{r stmodel}
stmodel <- stModel.define(
    smesh = smesh, ## spatial mesh
    tmesh = tmesh, ## temporal mesh
    model = '121', ## model, see the paper
    control.priors = list(
        prs = c(1, 0.1), ## P(spatial range < 1) = 0.1
        prt = c(5, 0), ## temporal range fixed to 5
        psigma = c(1, 0.1) ## P(sigma > 1) = 0.1
        )
    )
```

```{r stmodelloc, echo = FALSE}
stmodel <- stModel.define(
    smesh = smesh, ## spatial mesh
    tmesh = tmesh, ## temporal mesh
    model = '121', ## model, see the paper
    control.priors = list(
        prs = c(1, 0.1), ## P(spatial range < 1) = 0.1
        prt = c(5, 0), ## temporal range fixed to 5
        psigma = c(1, 0.1) ## P(sigma > 1) = 0.1
        ),
    useINLAprecomp = FALSE
    )
```

## Fit the model

Define a projector matrix from the spatial 
and temporal meshes to the data

```{r Aproj}
Aproj <- inla.spde.make.A(
    mesh = smesh,
    loc = cbind(dataf$s1, dataf$s2),
    group = dataf$time,
    group.mesh = tmesh
)
```

Create a 'fake' column to be used as index. 
in the `f()` term
```{r st}
dataf$st <- NA
```


Setting the likelihood precision (as fixed)
```{r precprior}
ctrl.lik <- list(
  hyper = list(
    prec = list(
      initial = 10, 
      fixed = TRUE)    
  )
)
```

Combine a 'fake' index column with `A.local`
```{r fmodel}
fmodel <- y ~ f(st, model = stmodel, A.local = Aproj)
```

Call the main `INLA` function:
```{r inla}
fit <- inla(
    formula = fmodel,
    data = dataf,
    control.family = ctrl.lik)
```

Posterior marginal summaries for fixed effect and 
the model parameters that were not fixed.
```{r pmdsummary}
fit$summary.fixed
fit$summary.hyperpar
```

## Using the **inlabru**

```{r inlabru}
library(inlabru)
```

Setting the observation (likelihood) model object
```{r obsmodel}
data_model <- bru_obs(
  formula = y ~ ., 
  family = "gaussian",
  control.family = ctrl.lik, 
  data = dataf)
```

Define the data model: the linear predictor terms
```{r linpred}
linpred <- ~ 1 +
    field(list(space = cbind(s1, s2), 
               time = time),
          model = stmodel)
```

Fitting
```{r fitting}
result <- bru(
  components = linpred,
  data_model)
```

Summary of the model parameters
```{r summary}
result$summary.fixed
result$summary.hyperpar
```
