---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# INLAspacetime

<!-- badges: start -->
[![CRAN Status](http://www.r-pkg.org/badges/version-last-release/INLAspacetime)](https://cran.r-project.org/package=INLAspacetime)
[![R build status](https://github.com/eliaskrainski/INLAspacetime/workflows/R-CMD-check/badge.svg)](https://github.com/eliaskrainski/INLAspacetime/actions)
<!--[![R code coverage status](https://github.com/eliaskrainski/INLAspacetime/workflows/test-coverage/badge.svg)](https://github.com/eliaskrainski/INLAspacetime/actions) -->
<!--[![lintr status](https://github.com/eliaskrainski/INLAspacetime/workflows/lint/badge.svg)](https://github.com/eliaskrainski/INLAspacetime/actions) -->
<!-- [![Codecov test coverage](https://codecov.io/gh/eliaskrainski/INLAspacetime/branch/master/graph/badge.svg)](https://app.codecov.io/gh/eliaskrainski/INLAspacetime?branch=master) -->
<!-- badges: end -->

This is a R package to implement certain spatial and spatio-temporal models
taking use to the `cgeneric` interface in the INLA package. 
This interface is a way to implement models by writing `C`
code to build the precision matrix compiling it so that 
INLA can use it internally.

## Installation

<!-- You can install the current [CRAN](https://CRAN.R-project.org) version of INLAspacetime: -->
<!-- ```{r cran-installation, eval = FALSE} -->
<!-- install.packages("INLAspacetime") -->
<!-- ``` -->

You can install the latest version of INLAspacetime from [GitHub](https://github.com/eliaskrainski/INLAspacetime) with
```{r ghinst, eval = FALSE}
## install.packages("remotes")
remotes::install_github("eliaskrainski/INLAspacetime")
```
<!-- or track the development version builds via [inlabru-org.r-universe.dev](https://inlabru-org.r-universe.dev/ui#builds): -->
<!-- ```{r universe-installation, eval = FALSE} -->
<!-- ## Enable universe(s) by inlabru-org -->
<!-- options(repos = c( -->
<!--   inlabruorg = "https://inlabru-org.r-universe.dev", -->
<!--   INLA = "https://inla.r-inla-download.org/R/testing", -->
<!--   CRAN = "https://cloud.r-project.org" -->
<!-- )) -->

<!-- ## Install it -->
<!-- install.packages("INLAspacetime") -->
<!-- ``` -->

# See the vignettes for examples

We will have  tutorials and examples at
[https://eliaskrainski.github.io/INLAspacetime/](https://eliaskrainski.github.io/INLAspacetime/)

## We have implemented 

1. some of the models presented in 
https://arxiv.org/abs/2006.04917

2. the barrier model proposed in 
https://doi.org/10.1016/j.spasta.2019.01.002

# Example

This is a basic example which fit a spacetime model for some fake data.
The model fitting using **inlabru** facilitates coding.

```{r fakedata}
n <- 5
dataf <- data.frame(
    s1   = runif(n, -1, 1),
    s2   = runif(n, -1, 1),
    time = runif(n, 1, 4),
    y    = rnorm(n, 0, 1))
str(dataf)
```

Loading the packages:
```{r library}
library(INLA)
library(INLAspacetime)
library(inlabru)
```

Define spatial and temporal discretization meshes
```{r meshes}
smesh <- inla.mesh.2d(
  loc = cbind(0,0), 
  max.edge = 5, 
  offset = 2)
tmesh <- inla.mesh.1d(
  loc = 0:5)
```

Define the spacetime model object to be used
```{r stmodel}
stmodel <- stModel.define(
    smesh = smesh, ## spatial mesh
    tmesh = tmesh, ## temporal mesh
    model = '121', ## model, see the paper
    control.priors = list(
        prs = c(1, 0.1), ## P(spatial range < 1) = 0.1
        prt = c(5, 0), ## fixed to 5
        psigma = c(1, 0.1) ## P(sigma > 1) = 0.1
        )
    )
```


Define the data model: the linear predictor terms
```{r dmodel}
linpred <- ~ 1 +
    field(list(space = cbind(s1, s2), 
               time = time),
          model = stmodel)
```

Setting the likelihood 
```{r precprior}
likeprec <- list(prec = list(
  initial = 10, fixed = FALSE))
datalike <- like(
  formula = y ~ ., 
  family = "gaussian",
  control.family = list(
    hyper = likeprec), ## TO DO: not going through bru
  data=dataf)
```

Fitting
```{r fitting}
result <- 
  bru(
    components = linpred,
    datalike,
    options = list(
      control.inla = list(
        int.strategy = "eb"
        ),
      verbose = !TRUE)
    )
```

Summary of the model parameters
```{r summary}
result$summary.fixed
result$summary.hyperpar
```
