---
title: "INLAspacetime: exploring with an example"
author: "Elias T Krainski"
date: "2023, Feb"
output: rmarkdown::html_vignette
bibliography: INLAspacetime.bib
vignette: >
  %\VignetteIndexEntry{INLAspacetime}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.width = 7,
  fig.height = 5
)
knitr::opts_chunk$set(echo = TRUE)
```

```{r cpackages}
library(ggplot2)
library(patchwork)
```

# Introduction 

In this vignette we illustrate how to fit some of the spacetime models
using the method proposed in @lindgren2023 
for the piemonte dataset analysed in @cameletti2013.

To perform this we will use the Bayesian paradigm with compuations done 
with the **INLA** package using the features provided by the 
**inlabru** package. 

```{r packages, results = "hide"}
library(INLA)
library(INLAspacetime)
library(inlabru)
```

We consider the following setting for **INLA**

```{r inlasetup}
inla.setOption(
  inla.mode='compact',
  smtp='pardiso',
  pardiso.license='~/.pardiso.lic')
```
```{r irem, echo = FALSE}
inla.setOption(
#  inla.call = "remote", 
  num.threads = "8:-1"
)
```

We will ask it to return the WAIC, DIC and CPO 
```{r ctrc}
ctrc <- list(
  waic=TRUE,
  dic=TRUE,
  cpo=TRUE)
```

# The dataset

Set the filenames

```{r filenames}
u0 <- paste0(
    'http://inla.r-inla-download.org/',
    'r-inla.org/case-studies/Cameletti2012/')
coofl <- 'coordinates.csv'
datafl <- 'Piemonte_data_byday.csv'
bordersfl <- 'Piemonte_borders.csv'
```

Download and read the borders file
```{r downloadporders}
### get the domain borders
if(!file.exists(bordersfl))
    download.file(paste0(u0, bordersfl), bordersfl)
dim(pborders <- read.csv(bordersfl))
```

Download and read the coordinates file
```{r coords}
### get the coordinates
if(!file.exists(coofl))
    download.file(paste0(u0, coofl), coofl)
dim(locs <- read.csv(coofl))
```

Download and read the dataset
```{r downloaddata}
### get the dataset
if(!file.exists(datafl))
    download.file(paste0(u0, datafl), datafl)
dim(pdata <- read.csv(datafl))
```

Inspect the dataset
```{r dinspect}
head(pdata)
```

Prepare the time to be used
```{r time}
range(pdata$Date <- as.Date(pdata$Date, '%d/%m/%y'))
pdata$time <- as.integer(difftime(
    pdata$Date, min(pdata$Date), units='days')) + 1
```

Standardize the covariates that will be used in the data analysis 
and define a dataset including the needed information 
where the outcome is the log of `PM10`, 
as used in @cameletti2013.
```{r covars}
### prepare the covariates
xnames <- c('A', 'WS', 'TEMP', 'HMIX', 'PREC', 'EMI')
xmean <- colMeans(pdata[, xnames])
xsd <- sapply(pdata[xnames], sd)

### prepare the data (st loc, scale covariates and log PM10)
dataf <- data.frame(pdata[c('UTMX', 'UTMY', 'time')],
                    scale(pdata[xnames], xmean, xsd),
                    y=log(pdata$PM10))
str(dataf)
```

# The model definition 

We consider the following linear mixed model for the outcome
\[ 
\mathbf{y} = \mathbf{W}\mathbf{\beta} + \mathbf{A}\mathbf{u} + \mathbf{e}
\]
where $\beta$ are fixed effects, or regression coefficients including the intercept, 
for the matrix of covariates $\mathbf{W}$,
$\mathbf{u}$ is the spatio-temporal random effect having the matrix
$\mathbf{A}$ the projector matrix from the discretized domain to the data.
The spatio-temporal random effect $\mathbf{u}$ is defined in a continuous 
spacetime domain being discretized considering meshes over time and space. 
The difference from @cameletti2013 is that we now use the models in 
@lindgren2023 for $\mathbf{u}$. 

Define a temporal mesh, with 1 knot per day.
```{r tmesh}
nt <- max(pdata$time)
tmesh <- inla.mesh.1d(1:nt, degree=1)
tmesh$n
```

Define a spatial mesh, the same used in @cameletti2013.
```{r smesh}
smesh <- inla.mesh.2d(
    cbind(locs[,2], locs[,3]),
    loc.domain=pborders,
    max.edge=c(50, 300),
    offset=c(10, 140),
    cutoff=5,
    min.angle=c(26, 21))
smesh$n
```

Visualize the spatial mesh, the border and the locations.
```{r smeshvis}
par(mfrow=c(1,1), mar=c(0,0,1,0))
plot(smesh, asp=1)
lines(pborders, lwd = 2, col = 'green4')
points(locs[, 2:3], pch = 12, col = "blue")
```

Add an overall integrate-to-zero constraint, 
which is not need but helps, in order to illustrate how it is done.
It has to consider the weights based on the mesh node volumes, 
that is, we consider the temporal and spatial ones as follows:
```{r stConstr}
tva <- inla.mesh.fem(tmesh)$c0@x
sva <- inla.mesh.fem(smesh)$va[,1]
stConstr <- list(
    A=matrix(kronecker(tva, sva), 1), e = 0)
```

Considering the **inlabru** package we start by defining the model components
```{r mcomp}
M <- ~ -1 + Intercept(1) + A + WS + TEMP + HMIX + PREC + EMI +
    field(list(space = cbind(UTMX, UTMY), time=time),
          model=stmodel, extraconstr=stConstr)
```

We se the prior for the likelihood precision 
considering a PC-prior, @pcprior, 
using a probabilistic statement 
P($\sigma_e > U_{\sigma_e}$) = $\alpha_{\sigma_e}$, 
using $U_{\sigma_e}$ = 10 and $\alpha_{\sigma_e} = 0.01$. 

```{r lprec}
lkprec <- list(
    prec=list(prior='pcprec', param=c(1, 0.1)))
```

The implementation of the spacetime model uses the `cgeneric` 
interface in **INLA**, see its documentation for details. 
Therefore we have a `C` code to mainly build the precision matrix and 
compute the model parameter priors and compiled as static library. 
We have this code included in the **INLAspacetime** package but it is also 
being copied to the **INLA** package and compiled with the same compilers 
in order to avoid possible mismatches. 
In order to use it, we have to define the matrices and vectors needed, 
including the prior parameter definitions.

The class of models in @lindgren2023 have the spatial range, 
temporal range and marginal standard deviation as parameters. 
We consider the PC-prior for these parameters defined from the 
probatility statements as 
P($r_s<U_{r_s}$)=$\alpha_{r_s}$, 
P($r_t<U_{r_t}$)=$\alpha_{r_t}$ and 
P($\sigma<U_{\sigma}$)=$\alpha_{\sigma}$.
We consider $U_{r_s}=30$, $U_{r_t}=1$ and $U_{\sigma}=10$.
$\alpha_{r_s}=\alpha_{r_t}=\alpha_{\sigma}=0.01$

The selection of one of the models in @lindgren2023 is by chosing the 
$\alpha_t$, $\alpha_s$ and $\alpha_e$ as integer numbers. 
We will start considering the model $\alpha_t=1$, $\alpha_s=0$ and $\alpha_t=2$, 
which is a model with separable spatio-temporal covariance, 
and then we fit some of the other models later.

```{r stmodeldef, echo = FALSE, eval = FALSE}
stmodel <- stModel.define(
    smesh, tmesh, model,
    control.priors = list(
        prs = c(50, 0.01),
        prt = c(1, 0.01),
        psigma = c(1, 0.01)))
```

We define an object with the needed 
use the function `stModel.define` 
where the model is selected considering the values for 
$\alpha_t$, $\alpha_s$ and $\alpha_e$ collapsed. 
```{r m102}
model <- "102"
<<stmodeldef>>
```

Initial values for the hyper-parameters help to fit the models in less computing time. 
It is also important to consider in the light that each dataset has its own parameter scale. 
For example, we have to consider that the spatial domain within a box of 
around $203.7$ by $266.5$ kilometers, which we already did when building the 
mesh and setti or $r_s$


```{r fit, echo = FALSE, eval = FALSE}
  bru(M,
      like(formula = y ~ .,
           family = "gaussian",
           control.family = list(
               hyper = lkprec),
           data = dataf),
      options = list(
          control.compute = ctrc))
```

```{r fitobj, echo = FALSE, eval = TRUE}
if(file.exists("fit102.rds")) {
  fit102 <- readRDS("fit102.rds")
} else {
  fit102 <- 
    <<fit>>
  saveRDS(fit102, file = "fit102.rds")
}
```

The code to fit the model through **inlabru** is
```{r fit102, eval = FALSE}
fit102 <- 
  <<fit>>
```

The computing time is
```{r cpu1}
fit102$cpu
```

Summary of the posterior marginal distributions for the fixed effects 
```{r sfixef}
fit102$summary.fixed[, c(1, 2, 3, 5)]
```

For the hyperparameters, we transform the posterior marginal distributions for the model hyperparameters
in the internal scale, 
$\log(1/\sigma^2_e)$, $\log(r_s)$, $\log(r_t)$ and $\log(\sigma)$,
to the user scale parametrization:
$\sigma_e$, $r_s$, $r_t$ and $\sigma$.
```{r thyper}
post.h <- list(
  sigma_e = inla.tmarginal(function(x) exp(-x/2), fit102$internal.marginals.hyperpar[[1]]),
  r_s = inla.tmarginal(function(x) exp(x), fit102$internal.marginals.hyperpar[[2]]),
  r_t = inla.tmarginal(function(x) exp(x), fit102$internal.marginals.hyperpar[[3]]),
  sigma = inla.tmarginal(function(x) exp(x), fit102$internal.marginals.hyperpar[[4]])
)
```

Then we compute and show the summary of it
```{r shyper}
shyper <- t(sapply(post.h, function(m) 
  unlist(inla.zmarginal(m, silent = TRUE))))
shyper[, c(1, 2, 3, 7)]
```

However, it is better to look at the posterior marginal itself, 
and we will visualize it later.

The model fitted in @cameletti2013 includes two more covariates and 
setup a model for discrete temporal domain where the temporal correlation 
is modeled as a first order autoregression with parameter $\rho$. 
In the fitted model here is defined considering continuous temporal domain 
with the range parameter $r_s$. 
However, the first order autocorrelation could be taken as 
$\rho = \exp(-2h/r_s)$, where $h$ is the temporal resolution used in 
the temporal mesh. As we used $h=1$ we can 
compare ou results with Table 3 in @cameletti2013 with

```{r compare1}
c(shyper[c(1, 4, 2), 1], 
  a=exp(-sqrt(8*0.5)/shyper[3, 1]))
```

# Comparing different models 

We now fit the model $121$ as well, 
using the same code, just building a new model object for $u$ 
as follows
```{r model121}
model <- "121"
<<stmodeldef>>
```

We create a list object and put each model fit 
as element on it, starting with the previously fitted model 
```{r fitl}
lfits <- vector("list", length = 2L) 
names(lfits) <- c("102", "121")
lfits[[1]] <- fit102
rm(fit102)
```

```{r omodels, eval = FALSE}
lfits[[2]] <- 
  <<fit>>
```
```{r modelsobj, echo = FALSE, eval = TRUE}
<<stmodeldef>>
flname <- paste0("fit", model, ".rds")
if(file.exists(flname)) {
   lfits[[2]] <- readRDS(flname)
} else {
  lfits[[2]] <-
     <<fit>>
  saveRDS(lfits[[2]], 
           file = flname)
}
```

The internal scale hyper-parameters posterior mode for each model
```{r modes}
sapply(lfits, function(r) r$mode$theta)
```

We will transform the posterior marginal distribution for the hyper-parameters 
as we did for the previous model 
```{r h2pmd}
posts.h2 <- lapply(1:2, function(m) vector("list", 4L))
for(m in 1:2) {
    posts.h2[[m]]$sigma_e =  
      data.frame(
        parameter = "sigma_e", 
        inla.tmarginal(
          function(x) exp(-x/2), 
          lfits[[m]]$internal.marginals.hyperpar[[1]]))
    for(p in 2:4) {
      posts.h2[[m]][[p]] <-   
      data.frame(
        parameter = c(NA, "r_s", "r_t", "sigma")[p], 
        inla.tmarginal(
          function(x) exp(x), 
          lfits[[m]]$internal.marginals.hyperpar[[p]])
      )
    }
}
```

Join these all to make visualization easier
```{r hjpmd}
posts.df <- rbind(
  data.frame(model = "102", do.call(rbind, posts.h2[[1]])),
  data.frame(model = "121", do.call(rbind, posts.h2[[2]]))
)

ggplot(posts.df) +
  geom_line(aes(x=x, y=y, group = model, color = model)) +
  ylab("Density") + xlab("") + 
  facet_wrap(~parameter, scales = "free")
```

The comparison of the model parameters of $\mathbf{u}$ 
for different models have to be done in light with the 
covariance functions as illustrated in @lindgren2023. 
The fitted $\sigma_e$ by the different models are comparable
and we can see that when considing model $121$ for $\mathbf{u}$, 
its posterior marginal are concentrated in values lower than 
when considering model $102$. 

We can also check fitting statistics such as DIC, WAIC, 
the negative of the log of the probability ordinates (LPO) 
and its cross-validated version (LCPO), 
summarized as the mean. 

```{r sfits}
t(sapply(lfits, function(r) {
  c(DIC = mean(r$dic$local.dic, na.rm = TRUE),
    WAIC = mean(r$waic$local.waic, na.rm = TRUE),
    LPO = -mean(r$po$po, na.rm = TRUE), 
    LCPO = -mean(r$cpo$cpo, na.rm = TRUE))
}))
```

# References
